DRAFT_Cleaning

### 2024-12-10 Draft_Cleaning.Rmd
continued to test and tune models, split data by river

```{r setup, include=FALSE}
library(readr)
library(tidyverse)
library(lubridate)
library(caret)
library(ggfortify)
library(ggformula)

@@

@@

data.frame(variable_contributions) |>
  gf_point(PC2 ~ PC1)

@@

predictor_matrix <- gage_survey_PCA |>
  dplyr::select(-VOLUMEDREDGED) %>% # Remove the response variable
  as.matrix() %>%
  scale() #centers and scales

@@

data.frame(data_in_PC_space) %>%
  gf_point(PC2 ~ PC1)

@@

###xgBoost
```{r}
library(xgboost)
gage_xgb <- gage_survey |>
  select(-lead_7, -lead_14,-DATE_START,
         -SURVEYJOBIDPK,-EXECYEAR,-DREDGE_TYPE,
         -DREDGINGPURPOSE)|>
    mutate_if(is.character, factor)

ctrl = trainControl(method = "cv", number = 10) 
fit_dredge = train(VOLUMEDREDGED ~ .,
                 data = gage_xgb,
                 method = "xgbTree",
                 tuneGrid = expand.grid(nrounds = c (15,25,50,100), max_depth = 1:10,
                                        eta = 0.3, gamma = 0,
                                        colsample_bytree = .8,
                                        min_child_weight = 1, subsample = 1),
                 verbosity = 0,
                 trControl = ctrl,
                 na.action = na.exclude)

varImp(fit_dredge)

@@

  ctrl = trainControl(method = "cv", number = 10)
  fit_dredge_DCV = train(VOLUMEDREDGED ~ .,
                       data = gage_xgb[in_train, ],
                       method = "xgbTree",
                       tuneGrid = expand.grid(nrounds = c (15,25,50,100), max_depth = 1:10,
                                        eta = 0.3, gamma = 0,
                                        colsample_bytree = 1,
                                        min_child_weight = 1, subsample = 1),
                       verbosity = 0,
                       trControl = ctrl,
                       na.action = na.exclude)

@@

tune_grid <- expand.grid(
  nrounds = c(50, 100, 150), 
  max_depth = c(3, 6, 9), 
  eta = c(0.1, 0.2, 0.3), 
  gamma = 0, 
  colsample_bytree = 0.8, 
  min_child_weight =1, 
  subsample = c(0.8, 1)
)

set.seed(123)
ctrl = trainControl(method = "cv", number = 10, savePredictions = "final")

fit_dredge2 = train(VOLUMEDREDGED ~ .,
                   data = gage_xgb,
                   method = "xgbTree",
                   tuneGrid = tune_grid,
                   trControl = ctrl,
                   na.action = na.exclude)

# Model performance
fit_dredge2

# Model evaluation
predictions <- fit_dredge2$pred  # Predictions from cross-validation
head(predictions)

# RMSE, R-Squared, etc.
cv_rmse <- sqrt(mean((predictions$obs - predictions$pred)^2))
cv_r2 <- cor(predictions$obs, predictions$pred)^2

cat("RMSE: ", cv_rmse, "\n")
cat("R-Squared: ", cv_r2, "\n")

@@

# I'll use RMSE to assess the model, because that's the metric I used to select the model.
# The model with the best RMSE may or may not be the one with the best MAE.
# RMSE
sqrt(mean((cv_pred - gage_xgb$VOLUMEDREDGED)^2)) # This is the most honest
                                         # assessment of accuracy

dredge_data_cleaning|>
  group_by(POOL)|>
  summarize(count())
```

>>>>

```{r setup, include=FALSE}
library(readr)
library(tidyverse)
library(lubridate)
library(caret)
library(ggfortify)
library(ggformula)
library(DiagrammeR)

@@

plot(dredge_PCA,  type="l")

autoplot(dredge_PCA)

@@

# data.frame(variable_contributions) |>
#   mutate(RIVER = gage_survey_cleaned$RIVER)
#   gf_point(PC2 ~ PC1)

@@

data.frame(variable_contributions) |>
  filter(abs(PC1) > 0.11)

predictor_matrix <- gage_survey_PCA |>
  dplyr::select(-VOLUMEDREDGED) |> # Remove the response variable
  as.matrix() |>
  scale() #centers and scales

@@

data_in_PC_space = data.frame(data_in_PC_space,
                              River = gage_survey_cleaned$RIVER,
                              Pool = gage_survey_cleaned$POOL,
                              DREDGE_TYPE = gage_survey_cleaned$DREDGINGPURPOSE,
                              Season = gage_survey_cleaned$Season)

data.frame(data_in_PC_space)|>
  gf_point(PC2 ~ PC1, col = ~ Season)

data.frame(data_in_PC_space)|>
  gf_point(PC2 ~ PC1, col = ~DREDGE_TYPE)

### Scree Plot
data_in_PC_space = predictor_matrix %*% variable_contributions
var_explained = apply(data_in_PC_space,2,var)
pve = var_explained/sum(var_explained)
pve

cumsum(pve)

var_df <- data.frame(pve, PC = 1:length(pve),
                     cum_var = cumsum(pve))

var_df|>
  gf_point(cum_var ~ PC) |>
  gf_line(cum_var ~ PC)

pca_scores <-dredge_PCA$x
pca_loadings <- dredge_PCA$rotation

@@

###xgBoost
```{r}
library(xgboost)
gage_xgb <- gage_survey_cleaned |>
  select(-lead_7, -lead_14,-DATE_START,
         -SURVEYJOBIDPK,-EXECYEAR,-DREDGE_TYPE,
         -DREDGINGPURPOSE)|>
    mutate_if(is.character, factor)

ctrl = trainControl(method = "cv", number = 10) 
fit_dredge = train(VOLUMEDREDGED ~ .,
                 data = gage_xgb,
                 method = "xgbTree",
               tuneGrid = expand.grid(nrounds = c (15,25,50,100), 
                                              max_depth = 1:10,
                                        eta =c(0.3), gamma = 0,
                                        colsample_bytree = 1,
                                        min_child_weight = 1, subsample = 1),
                 verbosity = 0,
                 trControl = ctrl,
                 na.action = na.exclude)

varImp(fit_dredge)
xgb.plot.tree(model = fit_dredge$finalModel, trees = 0)

@@

  ctrl = trainControl(method = "cv", number = 10)
  fit_dredge_DCV = train(VOLUMEDREDGED ~ .,
                       data = gage_xgb[in_train, ],
                       method = "xgbTree",
                       tuneGrid = expand.grid(nrounds = c (15,25,50,100), 
                                              max_depth = 1:10,
                                        eta =c(0.3), gamma = 0,
                                        colsample_bytree = 1,
                                        min_child_weight = 1, subsample = 1),
                       verbosity = 0,
                       trControl = ctrl,
                       na.action = na.exclude)

@@

@@

# I'll use RMSE to assess the model, because that's the metric I used to select the model.
# The model with the best RMSE may or may not be the one with the best MAE.
# RMSE
sqrt(mean((cv_pred - gage_xgb$VOLUMEDREDGED)^2)) # This is the most honest
                                         # assessment of accuracy
```

## Well that didnt work! Lets try modeling between the two river systems 
```{r}
combined_data <- bind_rows(survey_Data, dredge_data_cleaning)

Illinois_River <- gage_survey |>
  select(SURVEYJOBIDPK,EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED,RIVER,
         POOL, DREDGE_TYPE,UP_RIV_MIL, DN_RIV_MIL,lead_7,lead_14,BEAI2,HAVI2,
         HNYI2,IL03, IL03_Tail, IL04, IL04_Tail, IL05, IL05_Tail,IL06, IL06_Tail,
         IL07, IL07_Tail, IL08, IL08_Tail,KNGI2, MORI2, PIAI2) |>
  filter(RIVER == "Illinois_Waterway")

Miss_River <- gage_survey |>
  select(SURVEYJOBIDPK,EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED,RIVER,
         POOL, DREDGE_TYPE,UP_RIV_MIL, DN_RIV_MIL,lead_7,lead_14,BRLI4,CMMI4, DBQI4,
         FAII4, KHBI2,MI11,MI11_Tail,MI12,MI12_Tail,MI13,MI13_Tail,MI14,MI14_Tail,
         MI15,MI15_Tail,MI16,MI16_Tail,MI17,MI17_Tail,MI18,MI18_Tail,MI19,MI19_Tail,
         MI20,MI20_Tail,MI21,MI21_Tail,MI22,MI22_Tail, MUSI4,UINI2) |>
  filter(RIVER == "Mississippi_River")

Miss_full <- Miss_River|>  
  left_join(gage_data2, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
  left_join(gage_data2, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))

IWW_full <- Illinois_River|>  
  left_join(gage_data2, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
  left_join(gage_data2, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))

Miss_Cleaned <- Miss_full |>
  filter(complete.cases(Miss_full))

IWW_Cleaned <- IWW_full |>
  filter(complete.cases(IWW_full))

IWW_Cleaned <- IWW_Cleaned |>
  mutate("Season" = sapply(DATE_START,date_to_season))

Miss_Cleaned <- Miss_Cleaned |>
  mutate("Season" = sapply(DATE_START,date_to_season))

## xgboost split by river
#Miss
```{r}
Miss_xgb <-Miss_Cleaned |>
  select(-lead_7, -lead_14,-DATE_START,
         -SURVEYJOBIDPK,-EXECYEAR,-DREDGE_TYPE,
         -DREDGINGPURPOSE, -RIVER)|>
    mutate_if(is.character, factor)

ctrl = trainControl(method = "cv", number = 10) 
Miss_xgb_fit = train(VOLUMEDREDGED ~ .,
                 data = Miss_xgb,
                 method = "xgbTree",
               tuneGrid = expand.grid(nrounds = c(15,25,50,100), 
                                              max_depth = 1:10,
                                        eta =c(0.3), gamma = 0,
                                        colsample_bytree = 1,
                                        min_child_weight = 1, subsample = 1),
                 verbosity = 0,
                 trControl = ctrl,
                 na.action = na.exclude)

varImp(Miss_xgb_fit)
xgb.plot.tree(model = Miss_xgb_fit$finalModel, trees = 0)

Miss_xgb_fit

import_matrix = xgb.importance(model = Miss_xgb_fit$finalModel)
import_matrix
xgb.plot.importance(import_matrix)
```

#IWW
```{r}
IWW_xgb <-IWW_Cleaned |>
  select(-lead_7, -lead_14,-DATE_START,
         -SURVEYJOBIDPK,-EXECYEAR,-DREDGE_TYPE,
         -DREDGINGPURPOSE, -RIVER)|>
    mutate_if(is.character, factor)

ctrl = trainControl(method = "cv", number = 10) 
IWW_xgb_fit = train(VOLUMEDREDGED ~ .,
                 data = IWW_xgb,
                 method = "xgbTree",
               tuneGrid = expand.grid(nrounds = c(15,25,50,100), 
                                              max_depth = 1:10,
                                        eta =c(0.3), gamma = 0,
                                        colsample_bytree = 1,
                                        min_child_weight = 1, subsample = 1),
                 verbosity = 0,
                 trControl = ctrl,
                 na.action = na.exclude)

varImp(IWW_xgb_fit)
xgb.plot.tree(model = IWW_xgb_fit$finalModel, trees = 0)

IWW_xgb_fit

import_matrix = xgb.importance(model = IWW_xgb_fit$finalModel)
import_matrix
xgb.plot.importance(import_matrix)
```
